---
title: "GuitaR Tutorial"
author: "Artur Matysik"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

[**GuitaR**](https://arturmatysik.shinyapps.io/guitaR_app/) is a simple Shiny app to facilitate guitar soloing learning process, based on scales and intervals. 

URL: https://arturmatysik.shinyapps.io/guitaR_app/

In this tutorial I will walk through the design of GuitaR Shiny app with `tidyverse` metapackage. 

Topics include:
- Basic use of R functions
- Tidy-way of data structuring with `dplyr` 
- Creative use of `ggplot()`
- Simple implementation in R Shiny

## Generate fretboard

### Guitar setup

Start with guitar setup. Create list `guitar_setup` and store:

- `tuning`: vector of string names, e.g.: `c("E", "A", "D", "G", "B", "E")` will be the standard guitar tuning.
- `frets`: number of frets that will be later used for fretboard display.

```{r message=FALSE}
# setup
guitar_setup <- list(
  tuning = c("E", "A", "D", "G", "B", "E"),
  frets = 23
)
```

> Although we can provide separately number of strings that guitar uses, it will be redundant, because string can't exist without tuning! If necessary, number of strings can be easily calculated (e.g. `length(guitar_setup$tuning)`).

### Fretboard layout

#### Simple fretboard

From the setup we know the number of frets as well as number and names of strings. That it all we need to draw an empty fretboard. 

- Use `geom_vline()` to draw frets and nut (with significantly thicker line)
- Use `geom_hline()` to draw strings
- Use `scale_y_continuous()` to label strings
  - set `breaks` to `1:length(guitar_setup$tuning)`
  - set `labels` to `guitar_setup$tuning`
- Use `scale_x_continuous()` to label strings 
  - `breaks` and `labels` equal to `1:guitar_setup$frets`
- Remove unwanted margin with `expand()`
- Use `theme_minimal()` for minimalist design
- Use `theme()` attributes to 
  - remove grid `panel.grid.minor = element_blank()`
  - remove axis labels `axis.title=element_blank()`

```{r message=FALSE, fig.dim = c(10, 2)}
library(tidyverse)
ggplot(NULL) +
  # plot frets
  geom_vline(xintercept = 1:guitar_setup$frets, alpha = 0.3) +
  # plot nut
  geom_vline(xintercept = 0, alpha = 1, size = 2) +
  # plot strings
  geom_hline(yintercept = 1:length(guitar_setup$tuning)) + 
  # label strings
  scale_y_continuous(
    breaks = 1:length(guitar_setup$tuning),
    labels = guitar_setup$tuning,
    expand = c(0.07, 0.07)
  ) + 
  # label frets
  scale_x_continuous(
    breaks = 1:guitar_setup$frets,
    labels = 1:guitar_setup$frets,
    expand = c(.005, .005)
  ) + 
  theme_minimal() + 
  theme(
    # remove unwanted grids
    panel.grid.minor = element_blank(),
    # remove axis labels
    axis.title=element_blank())
```

That looks good already! But lets make it even better.

#### Fret distance

As we know, real guitar frets are not evenly spread. Their distance is bigger close to the guitar neck, and smaller towards the bridge. Without going into details, the following formula can help in calculating fret distance: `1 - exp(-k * n)` where `n` is the fret number and `k` is a constant of `k = 5.71584144995393e-2`. We can use this formula to generate function, that takes fret number and returns its relative position:

```{r}
calc_fret_distance <- function(n, s = 1, k = 5.71584144995393e-2) {
  s * (1 - exp(-k * n))
}

# position of nut
calc_fret_distance(0)

# position of 1st fret
calc_fret_distance(1)

# position of 10th fret
calc_fret_distance(10)
```

Having this convenient way of fret position calculation, we can create simple data frame `fretboard_layout`, to hold fret number (`fret_num`) and fret position `fret_pos`):

```{r message=TRUE}
library(tidyverse)
fretboard_layout <- data.frame(
  fret_num = 0:guitar_setup$frets,
  fret_pos = calc_fret_distance(0:guitar_setup$frets)
)
head(fretboard_layout)
```

#### Fretboard markers

Another missing things are the markers. Lets add them to the `fretboard_layout` data frame.
For standard guitar markers are located at frets 3, 5, 7, 9, 12, 15, 17, 19, 21. 

Typically, there are two types:

- single markers (`marker = "single"`), located at frets 3, 5, 7, 9, 15, 17, 19, 21 (`fret_num = c(3, 5, 7, 9, 15, 17, 19, 21)`)
- double marker (`marker = "double"`) at fret 12 (`fret_num = c(12)`)

Their position however is not exactly where the fret is, but rather between two frets. 
For example: marker at 5th fret will lie between 4th and 5th fret. 

First, create function `calc_fret_marker_position`, that will take fret number and vector of positions (`fretboard$fret_pos`) to calculate mean position between two frets. 

```{r}
calc_fret_marker_position <- function(n, pos) {
  mean(c(pos[n], pos[n + 1]))
}
```

Use `Vectorize()` to enable vector input:

```{r}
calc_fret_marker_position_v <-
  Vectorize(calc_fret_marker_position, "n")
```

Then, use vectorized function to calculate `fretboard_markers` data frame, holding marker X positions (`marker_posX`). 

As for the Y position (`marker_posY`):

- Single markers are place in the middle (`max(length(guitar_setup$tuning) + 1) / 2`)
- For double markers calculate middle point as for single, then double and separate +1 up or down by adding `+ c(-1, 1)` to the middle position. Note, that by multiplying `c(-1, 1)` by some value we can control distance between the two markers!

```{r}
fretboard_markers <- bind_rows(
  # fret vs marker
  data.frame(
    marker = "single",
    fret_num = c(3, 5, 7, 9, 15, 17, 19, 21),
    marker_posY = max(length(guitar_setup$tuning) + 1) / 2
  ),
  data.frame(
    marker = "double",
    fret_num = c(12),
    # c(-1, 1) * 1: * 1 can be a distance parameter here
    marker_posY = max(length(guitar_setup$tuning) + 1) / 2 + c(-1, 1) * 1
  )
) %>%
  # join with layout
  right_join(fretboard_layout, by = "fret_num") %>%
  arrange(fret_num) %>%
  # calc marker position X (fret)
  mutate(marker_posX = calc_fret_marker_position_v(fret_num, fret_pos)) %>%
  filter(!is.na(marker)) %>%
  select(fret_num, marker, marker_posX, marker_posY)
```

Finally, join `fretboard_markers` with `fretboard_layout` to get the final `fretboard`, that we will use for plotting.

```{r}
fretboard <- left_join(fretboard_layout, fretboard_markers, by = "fret_num")

head(fretboard)
```

#### Fancy fretboard

We can combine all above functions into one `calc_fretboard()`, that will take `guitar_setup` and return the `fretboard`:

```{r}
# design fretboard
calc_fretboard <- function(guitar_setup) {
  # calculate layout
  calc_fret_distance <- function(n, s = 1, k = 5.71584144995393e-2) {
    s * (1 - exp(-k * n))
  }

  fretboard_layout <- data.frame(
    fret_pos = calc_fret_distance(0:guitar_setup$frets),
    fret_num = 0:guitar_setup$frets
  ) %>% select(fret_num, fret_pos)

  # calculate markers
  calc_fret_marker_position <- function(n, pos) {
    mean(c(pos[n], pos[n+1]))
  }
  calc_fret_marker_position_v <- Vectorize(calc_fret_marker_position, "n")

  fretboard_markers <- bind_rows(
    # fret vs marker
    data.frame(marker = "single",
               fret_num = c(3, 5, 7, 9, 15, 17, 19, 21),
               marker_posY = max(length(guitar_setup$tuning) + 1) / 2),
    data.frame(marker = "double",
               fret_num = c(12),
               # c(-1, 1) * 1: * 1 can be a distance parameter here
               marker_posY = max(length(guitar_setup$tuning) + 1) / 2 + c(-1, 1) * 1)) %>%
    # join with layout
    right_join(fretboard_layout, by = "fret_num") %>%
    arrange(fret_num) %>%
    # calc marker position X (fret)
    mutate(marker_posX = calc_fret_marker_position_v(fret_num, fret_pos)) %>%
    filter(!is.na(marker)) %>%
    select(fret_num, marker, marker_posX, marker_posY)

  left_join(fretboard_layout, fretboard_markers, by = "fret_num")
}

fretboard <- calc_fretboard(guitar_setup)
```

#### Fancy fretboard - plot

Use `fretboard` for plotting the fretboard:

```{r, fig.dim = c(10, 2)}
# set drawing parameters
note_label_size <- 4
note_point_size <- 8

# plot empty fretboard
fr <- ggplot(NULL) +
  # plot frets
  geom_vline(data = fretboard, aes(xintercept = fret_pos), alpha = 0.3) +
  geom_vline(xintercept = 0, alpha = 1, size = 2) +
  # plot strings
  geom_hline(
    yintercept = 1:length(guitar_setup$tuning),
    # string thickness
    size = 1 - seq(
      from = 0,
      to = 0.5,
      length.out = length(guitar_setup$tuning)
    )
  ) +
  # plot single markers
  geom_point(
    data = fretboard %>% filter(marker == "single"),
    aes(x = marker_posX, y = marker_posY),
    size = 5,
    alpha = 0.1
  ) +
  # plot double markers
  geom_point(
    data = fretboard %>% filter(marker == "double"),
    aes(x = marker_posX, y = marker_posY),
    size = 5,
    alpha = 0.1
  ) +
  # label strings
  scale_y_continuous(
    breaks = 1:length(guitar_setup$tuning),
    labels = guitar_setup$tuning,
    expand = c(0.07, 0.07)
  ) +
  # label frets
  scale_x_continuous(
    breaks = fretboard$fret_pos,
    labels = fretboard$fret_num,
    expand = c(.005, .005)
  ) +
  # plot labels
  # labs(x = "fret", y = "string") +
  # theme
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        axis.title=element_blank()
        )

fr
```

